Forecasting Gamma-ray Bursts using Gravitational Waves

### 2018-03-13 Paper/Manuscript.tex
2 lines to intro

\section{Introduction}\label{Sec:Intro}
We reserve the symbol $f$ to denote the GW frequency of the dominant quadrupole mode, which is twice the orbital frequency $f_K$. 
Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol for quantities which we can compute exactly, but list finite digits and $\approx$ for quantities whose values we know approximately.
Overdots denote time derivative with respect to detector time, e.g., $\dot{E} =dE/dt$.

@@

\begin{table}[h]
\centering
\begin{tabular}{l|ccccccc}
%toprule
\hline
$D\,$(Mpc) & \multicolumn{3}{c}{ET-B} &  & \multicolumn{3}{c}{ET-C}\T\B\\
\hline
{}& $\bar{f}\,$(Hz) & \ $T_\text{AW}$(sec) \ & $\bar{\rho}_{F}$ &{} & $\bar{f}\,$(Hz) & \ $T_\text{AW}$(sec)\ & $\bar{\rho}_{F}$\T\B\\
%midrule

100 & 6.7237 &  2818 & 306.0 &{\qquad} & 3.2692 & 19212 & 365.4\T\\
200 & 11.2104 & 698.6 & 152.4 &{\qquad} & 4.1028 & 10138 & 182.1 \\
400 & 18.2234 & 179.8 & 75.74 &{\qquad} & 5.0607 & 5437 & 90.48\\
1000 & 41.3537 &17.2 & 29.79& \qquad & 6.7553 & 2138 & 35.60  \\
\hline
%bottomrule
\end{tabular}
\caption{Forecasting capabilities of Einstein Telescope summarized. We can expect over \emph{five} hours of early warning for nearby sources. 
ET-B and ET-C refer the different configurations shown in Fig.~\ref{fig:ETB2030}. For the advance warning times, we only present the result of the more accurate 3.5PN computation. $\bar{f}$ is the threshold frequency at which
ET-B/C accumulate SNR of 15 which we take to be our detection criterion. Note that both $T_\text{AW}$ and $\bar\rho_F$ are larger for ET-C
due to its improved sensitivity in the $f\lesssim 30\,$Hz regime compared to ET-B as it is clear in Fig.~\ref{fig:ETB2030}.
These results and those of Table.~\ref{table:LIGO2020} are summarized in Fig.~\ref{fig:summary}.}\label{table:ET}
\end{table}
%
%
As the results of Table~\ref{table:ET} indicate, ET-B will provide $\sim 10 - 50\,$ minutes of warning time for BNS inspirals at $D\lesssim 200\,$Mpc.
Instead of speculating whether or not this is enough to forecast GRBs, we turn our attention to ET-C's forecasting capabilities.
As can be seen from Fig.~\ref{fig:ETB2030}, ET-C's increased sensitivity in the $f\lesssim 30\,$Hz regime will significantly lengthen the advance warning times.
This is best highlighted by the sixth colum of Table~\ref{table:ET}, 
where we show that ET-C will provide up to five hours of early warning time for nearby ($D\lesssim 100\,$Mpc) neutron star mergers.
Moreover, BNS inspirals as far as 200\,Mpc (the ALV network's horizon distance) will persist in the ET-C sensitivity band for a few hours after their initial detection.
Current BNS merger rate, inferred from A-LIGO's O1-O2 periods, implies that $\sim 40$ to $\sim600$ sources within a volume of $(200\,\text{Mpc})^3$ will be detected yearly by ET.
Given Moore's law and new search approaches based on deep learning (\cite{Gabbard:2017lja}) we speculate
that the advances in the detection, localization and early-warning algorithms in the next decade will reduce the required early
warning time to less than an hour; thus we set $T_\text{AW} = 1\,\text{hour}$ and obtain corresponding BNS horizon distances of $D_H=87,613\,$Mpc ($z\approx 0.02,0.127$) for ET-B and ET-C, respectively.
In other words, all inspiralling BNSs closer than $D_H$ will give us more than an hour's leeway before the GRB (assuming detection SNR of 15 as always).
The volume set by $D=D_H$ translates to a BNS merger rate of $\mathcal{O}(1)$ for ET-B and
$\approx 355^{+730}_{-280}\,\text{year}^{-1}$ for ET-C. Additionally, each source within $D_H$ will accumulate $\bar\rho_F \gtrsim 420, 58$ over its inspiral
in the ET-B, ET-C bands, respectively. These are summarized in Table~\ref{table:horizon}.

@@

So we will simply assume that such lightweight, high-spin BHs exist in BH-NS binaries and compute strains and SNRs accordingly.
A quick calculation shows that these systems yield $T_\text{AW} \lesssim 30\,$seconds in the ALV network of 2020s. %in Table \ref{table:LIGO2020}is now less than $ 40\,$ seconds. 
Therefore, we simply focus on the capabilities of ET which we summarize in Table~\ref{table:ET_BH_NS} below.
%
%
%
\begin{table}[h]
\centering
\begin{tabular}{l|ccccc}
%toprule
\hline
$D\,$(Mpc) & \multicolumn{2}{c}{ET-B} &  & \multicolumn{2}{c}{ET-C}\T\B\\
\hline
{}&  \ \hspace{6mm}$T_\text{AW}$ \hspace{8mm} & $\bar{\rho}_{F}$ &{}  & \hspace{6mm}$T_\text{AW}$ \hspace{8mm} & $\bar{\rho}_{F}$\T\B\\
%midrule

100 &   28.2 minutes & 503 &{\qquad} &  2.15 hours & 601\T\\
200 & 8.2 minutes & 251  &{\qquad} & 1.1 hours & 300 \\
400 &  2 minutes & 124 &{\qquad} & 34 minutes & 148\\
1000 & 14.3 seconds & 49 &{\qquad} & 14 minutes& 58.5\\
\hline
%bottomrule
\end{tabular}
\caption{Same as Table~\ref{table:ET}, but now for a $5 M_\odot- 1.4 M_\odot$ black hole-neutron star binary.
The SNR is even higher than before, however the advance warning times have decreased significantly.}\label{table:ET_BH_NS}
\end{table}
%
%
\\
\noindent
From the table, we can see that the horizon distance of ET-C for these binaries with $T_\text{AW} \approx 1\,$hour is roughly 200\,Mpc.
Given that the LIGO O1 merger rate for these systems is $ R < 3600\,$Gpc$^{-3}$yr$^{-1}$ \cite{Abbott:2016ymx},
there could be 28.8 such mergers per year within ET-C's horizon distance. We must emphasize that this should be viewed
as an upper limit as low-mass, high-spin black holes such as ours above are expected to be rare.
Therefore, we realistically expect ET-C to forecast $\sim \mathcal{O}(1)$ tidal disruption event per year.
A better estimation would need to incorporate the percentage of such black holes from population studies and Monte-Carlo simulations,
however, all of these come with rather large error bars {\bf CITE}.
%and that no TD occurs for $a=0$. This is roughly consistent with Fig.5 of Ref.~\cite{Hanna:2008um}
%Choosing lower mass implies reduced SNR since $\rho \propto M_c^{5/3} \propto (1+Q)^{1/3}/Q$,
%but even with $m_2=5 M_\odot$, a BH-NS binary is ``louder'' than a BNS system at the same distance. 
%and as we showed in Sec.~\ref{sec:results}, low SNR is not going to be a problem with ET sensitivities out to $D \sim 1\,$Gpc.
%
%It is straightforward to show that this mass bound is only consistent with spinning black holes by substituting the naive Schwarzschild $r_\text{ISCO}$ value 
%into Eq~.(\ref{eq:TD_cond1}) along with $c_R=1.7$ {\bf CITE} which yield $M_\text{BH} \lesssim 4.1 M_\odot$ violating the above BH mass bound.
%ISCO location from NR 
%\be
%\Omega_\text{ISCO} M = 0.068 \left[ 1- \f{0.444}{Q^{1/4}}(1-3.54 C^{1/3})\right] 
%\ee

>>>>

\section{Introduction}\label{Sec:Intro}
The 6.9 gravitational-wave merger events detected by the ALV network have firmly established gravitational-wave astronomy
as observational astronomy. Though the first event of 14 September 2015 will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 event involving the inspiral and merger of a binary neutron star system with a concordant gamma-ray burst detected 1.4 seconds after the coalescence.

We reserve the symbol $f$ to denote the GW frequency of the dominant quadrupole mode, which is twice the orbital frequency $f_K$. 
Relatedly, we define the following angular frequencies: $\omega = 2\pi f, \Omega=2\pi f_K$ with $\omega=2\Omega$.
We employ the $\simeq$ symbol for quantities which we can compute exactly, but list finite digits and $\approx$ for quantities whose values we know approximately.
Overdots denote time derivative with respect to detector time, e.g., $\dot{E} =dE/dt$.

@@

\begin{table}[h]
\centering
\begin{tabular}{l|ccccccc}
%toprule
\hline
$D\,$(Mpc) & \multicolumn{3}{c}{ET-B} &  & \multicolumn{3}{c}{ET-C}\T\B\\
\hline
{}& $\bar{f}\,$(Hz) & \ \hspace{7mm} $T_\text{AW}$ \ \hspace{7mm} & $\bar{\rho}_{F}$ &{} & $\bar{f}\,$(Hz) & \ \hspace{5mm} $T_\text{AW}\hspace{12mm}$& $\bar{\rho}_{F}$\T\B\\
%midrule
100 & $\approx\,$6.72 &  47.0\,minutes & 306.0 &{\qquad} & $\approx\,$3.27 & 5.34\,hours\ & 365.4\T\\
200 & $\approx\,$11.2 & 11.6\,minutes & 152.4 &{\qquad} & $\approx\,$4.10 & 2.87\,hours\ & 182.1 \\
400 & $\approx\,$18.2 & 3.00\,minutes & 75.74 &{\qquad} & $\approx\,$5.06 & 1.51\,hours\ & 90.48\\
1000 & $\approx\,$41.3 &17.2\,seconds & 29.79& \qquad &    $\approx\,$6.76 & 35.6\,minutes & 35.60  \\
% OLD version
%100 & 6.7237 &  2818 & 306.0 &{\qquad} & 3.2692 & 19212 & 365.4\T\\
%200 & 11.2104 & 698.6 & 152.4 &{\qquad} & 4.1028 & 10138 & 182.1 \\
%400 & 18.2234 & 179.8 & 75.74 &{\qquad} & 5.0607 & 5437 & 90.48\\
%1000 & 41.3537 &17.2 & 29.79& \qquad & 6.7553 & 2138 & 35.60  \\
\hline
%bottomrule
\end{tabular}
\caption{Forecasting capabilities of Einstein Telescope summarized. We can expect over \emph{five} hours of early warning for nearby sources. 
ET-B and ET-C refer the different configurations shown in Fig.~\ref{fig:ETB2030}. For the advance warning times, we only present the result of the more accurate 3.5PN computation. $\bar{f}$ is the threshold frequency at which
ET-B/C accumulate SNR of 15 which we take to be our detection criterion. Note that both $T_\text{AW}$ and $\bar\rho_F$ are larger for ET-C
due to its improved sensitivity in the $f\lesssim 30\,$Hz regime compared to ET-B as it is clear in Fig.~\ref{fig:ETB2030}.
These results and those of Table.~\ref{table:LIGO2020} are summarized in Fig.~\ref{fig:summary}.}\label{table:ET}
\end{table}
%
%
As the results of Table~\ref{table:ET} indicate, ET-B will provide $\sim 10 - 50\,$ minutes of warning time for BNS inspirals at $D\lesssim 200\,$Mpc.
Instead of speculating whether or not this is enough to forecast GRBs, we turn our attention to ET-C's forecasting capabilities.
As can be seen from Fig.~\ref{fig:ETB2030}, ET-C's increased sensitivity in the $f\lesssim 30\,$Hz regime will significantly lengthen the advance warning times.
This is best highlighted by the sixth colum of Table~\ref{table:ET}, 
where we show that ET-C will provide up to five hours of early warning time for nearby ($D\lesssim 100\,$Mpc) neutron star mergers.
Moreover, BNS inspirals as far as 200\,Mpc (the ALV network's horizon distance) will persist in the ET-C sensitivity band for a few hours after their initial detection.
Current BNS merger rate, inferred from A-LIGO's O1-O2 periods, implies that $\sim 40$ to $\sim600$ sources within a volume of $(200\,\text{Mpc})^3$ will be detected yearly by ET.
Given Moore's law and new search approaches based on deep learning (\cite{Gabbard:2017lja}) we speculate
that the advances in the detection, localization and early-warning algorithms in the next decade will reduce the required early
warning time to less than an hour; thus we set $T_\text{AW} = 1\,\text{hour}$ and obtain corresponding BNS horizon distances of $D_H=87,613\,$Mpc ($z\approx 0.02,0.127$) for ET-B and ET-C, respectively.
In other words, all inspiralling BNSs closer than $D_H$ will give us more than an hour's leeway before the GRB (assuming detection SNR of 15 as always).
The volume set by $D=D_H$ translates to a BNS merger rate of $\mathcal{O}(1)$ for ET-B and
$\approx 355^{+730}_{-280}\,\text{year}^{-1}$ for ET-C. Additionally, each source within $D_H$ will accumulate $\bar\rho_F \gtrsim 420, 58$ over its inspiral
in the ET-B, ET-C bands, respectively. These are summarized in Table~\ref{table:horizon}.

@@

So we will simply assume that such lightweight, high-spin BHs exist in BH-NS binaries and compute strains and SNRs accordingly.
A quick calculation shows that these systems yield $T_\text{AW} \lesssim 30\,$seconds in the ALV network of 2020s. %in Table \ref{table:LIGO2020}is now less than $ 40\,$ seconds. 
Therefore, we simply focus on the capabilities of ET which we summarize in Table~\ref{table:ET_BH_NS} below.
%
%
%
\begin{table}[h]
\centering
\begin{tabular}{l|ccccc}
%toprule
\hline
$D\,$(Mpc) & \multicolumn{2}{c}{ET-B} &  & \multicolumn{2}{c}{ET-C}\T\B\\
\hline
{}&  \ \hspace{6mm}$T_\text{AW}$ \hspace{8mm} & $\bar{\rho}_{F}$ &{}  & $\hspace{6mm}T_\text{AW} \hspace{10mm}$ & $\bar{\rho}_{F}$\T\B\\
%midrule

100 &   28.2\,minutes & 503 &{\qquad} &  2.15\,hours & 601\T\\
200 & 8.21\,minutes & 251  &{\qquad} & 1.10\,hours & 300 \\
400 &  2.00\,minutes & 124 &{\qquad} & 34.2\,minutes & 148\\
1000 & 14.3\,seconds & 49 &{\qquad} & 13.9\,minutes& 58.5\\
\hline
%bottomrule
\end{tabular}
\caption{Same as Table~\ref{table:ET}, but now for a $5 M_\odot- 1.4 M_\odot$ black hole-neutron star binary.
The SNR is even higher than before, however the advance warning times have decreased significantly.}\label{table:ET_BH_NS}
\end{table}
%
%
\\
\noindent
From the table, we can see that the horizon distance of ET-C for these binaries with $T_\text{AW} \approx 1\,$hour is roughly 200\,Mpc.
Given that the LIGO O1 merger rate for these systems is $ R < 3600\,$Gpc$^{-3}$yr$^{-1}$ \cite{Abbott:2016ymx},
there could be 28.8 such mergers per year within ET-C's horizon distance. We must emphasize that this should be viewed
as an upper limit as low-mass, high-spin black holes such as ours above are expected to be rare.
Therefore, we realistically expect ET-C to forecast $\sim \mathcal{O}(1)$ tidal disruption event per year.
A better estimation would need to incorporate the percentage of such black holes from population studies and Monte-Carlo simulations,
however, all of these come with rather large error bars {\bf CITE}.
%and that no TD occurs for $a=0$. This is roughly consistent with Fig.5 of Ref.~\cite{Hanna:2008um}
%Choosing lower mass implies reduced SNR since $\rho \propto M_c^{5/3} \propto (1+Q)^{1/3}/Q$,
%but even with $m_2=5 M_\odot$, a BH-NS binary is ``louder'' than a BNS system at the same distance. 
%and as we showed in Sec.~\ref{sec:results}, low SNR is not going to be a problem with ET sensitivities out to $D \sim 1\,$Gpc.
%
%It is straightforward to show that this mass bound is only consistent with spinning black holes by substituting the naive Schwarzschild $r_\text{ISCO}$ value 
%into Eq~.(\ref{eq:TD_cond1}) along with $c_R=1.7$ {\bf CITE} which yield $M_\text{BH} \lesssim 4.1 M_\odot$ violating the above BH mass bound.
%ISCO location from NR 
%\be
%\Omega_\text{ISCO} M = 0.068 \left[ 1- \f{0.444}{Q^{1/4}}(1-3.54 C^{1/3})\right] 
%\ee
